---
title: "Aggregating Distance Metrics"
author: "Kent Johnson"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    df_print: kable
    toc: true
vignette: >
  %\VignetteIndexEntry{Aggregating Distance Metrics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo=FALSE,include=FALSE,message=FALSE}
suppressPackageStartupMessages(library(tidyverse))
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

<style type="text/css">
img { 
  border: none;
}
table {
    width: auto !important;
}
</style>

This vignette uses the `phenoptrExamples` sample data
and functions from the [tidyverse](http://tidyverse.org/) to demonstrate
reading and processing cell seg data from multiple fields and samples.

## Read multiple data files

Use `list_cell_seg_files` and `purrr::map_df` to read all
cell seg data files in a single directory into a single `data_frame`.
The result is
similar to reading an inForm merge table.

### Find all cell seg data files in a directory

`list_cell_seg_files` takes a directory path as an argument and 
returns a list of paths to all the `cell_seg_data.txt`
files in a directory.

```{r}
library(phenoptr)
library(tidyverse)
base_path = system.file("extdata", "samples", package = "phenoptrExamples")
paths = list_cell_seg_files(base_path)
length(paths)
paths[1]
```

### Read and combine files

`purrr::map_df` applies `read_cell_seg_data` to each path in `paths`. The
`data_frame`s returned from each
call to `read_cell_seg_data` are combined row-wise to create a single
merged `data_frame`. 

```{r}
csd = purrr::map_df(paths, read_cell_seg_data)
dim(csd)
```

Using `table` is one way to summarize the data by `Sample Name` or `Slide ID`.
This data comes from nine fields taken of three slides.

```{r}
table(csd$`Sample Name`, csd$Phenotype) %>% addmargins(2, list(Total=sum))
table(csd$`Slide ID`, csd$Phenotype) %>% addmargins(2, list(Total=sum))
```

## Compute on merged data

Use `dplyr:group_by` and `dplyr::summarize`
to compute summary statistics for all fields
in a slide. For finer grouping, use multiple arguments to `group_by`. 
Use `dplyr::filter` to select a particular phenotype tissue category.

This example computes the mean PDL1 expression for `CD68+` and `CK+` cells
in `Tumor`, with the mean computed per `Slide ID`

```{r}
csd %>% 
  filter(`Tissue Category`=='Tumor', Phenotype %in% c('CD68+', 'CK+')) %>% 
  group_by(`Slide ID`, Phenotype) %>% 
  summarize(Mean_PDL1=mean(`Entire Cell PDL1 (Opal 520) Mean`))
```

### Add distance columns

Nearest-neighbor distances must be computed per-sample because the X/Y
coordinates reported in cell seg data files are all relative to the top-left
of the sample.

Use `dplyr::group_by` to aggregate across subsets of a full
data set. In this case, we want to group by `Sample Name`. Within each
group, use `dplyr::do` to call `find_nearest_distance` to compute the distance
columns and `dplyr::bind_cols` to combine them with the original data.

```{r eval=FALSE}
# Use the same list of phenotypes for each sample
phenos = unique(csd$Phenotype)
csd = csd %>%
  group_by(`Sample Name`) %>%
  do(bind_cols(., find_nearest_distance(., phenos)))
dim(csd)
tail(names(csd), 5)
```
```{r echo=FALSE}
# Use cached values
distances = read_csv('nearest_distance.csv')
csd = bind_cols(csd, distances)
dim(csd)
tail(names(csd), 5)
```


### Average distance per sample

The next example uses `group_by`, `filter` and `summarize` again to
compute the average distance from a tumor cell
(`CK+`) to the nearest macrophage (`CD68+`), with the averages computed
per `Slide ID`.

```{r}
csd %>% group_by(`Slide ID`) %>% 
  filter(Phenotype=='CK+') %>% # Only tumor cells
  summarize(mean_dist_to_CD68=round(mean(`Distance to CD68+`), 2))
```

## Aggregate `count_within` across samples

### Compute counts and averages

Use `count_within_batch` to count cells within a radius for 
multiple tissue categories, phenotypes and fields. 
This example counts `CK+` cells with a `CD8+` cell within 10 or 25 microns,
and `CK+` cells with a `CD68+` cell within 10 or 25 microns.
`dplyr::glimpse` gives a compact summary of the data.

```{r eval=FALSE}
base_path = system.file("extdata", "samples", package = "phenoptrExamples")
pairs = list(c('CK+', 'CD8+'),
             c('CK+', 'CD68+'))
radius = c(10, 25)
counts = count_within_batch(base_path, pairs, radius, verbose=FALSE) %>% 
  select(-source, -category) # Remove unneeded columns

glimpse(counts)
```
```{r echo=FALSE}
# count_with_batch is slow, we cheat by caching the result
counts = read_csv('count_within_batch.csv')
glimpse(counts)
```

### Aggregate counts and means per sample

Aggregating `from_count`, `to_count` and `from_with`
across samples is straighforward, it only requires
simple sums. Aggregating `within_mean` requires computing the underlying 
count of cells within the radius, summing, and computing a new mean.

(Note: the value of `from_count * within_mean` is not reported by
`count_with` because it may count cells multiple times.)

```{r}
counts_per_sample = counts %>% group_by(slide_id, from, to, radius) %>% 
    summarize(from_count=sum(from_count),
              to_count=sum(to_count),
              from_with=sum(from_with),
              within=sum(from_count*within_mean),
              within_mean=within/from_count) %>%
  ungroup %>% select(-within)
counts_per_sample
```

