---
title: "Aggregating touch counts"
author: "Kent Johnson"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    df_print: kable
    toc: true
vignette: >
  %\VignetteIndexEntry{Aggregating touch counts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo=FALSE,include=FALSE,message=FALSE}
suppressPackageStartupMessages(library(tidyverse))
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

The `phenoptr::count_touching_cells` function finds, counts and 
visualizes touching cells of multiple phenotypes in a single field.

This vignette gives an example of counting touches in multiple fields
and aggregating across slides. This vignette does not create visualizations
of the touching cells; the comments show how to change this.

----

## Count touching cells in multiple fields

This example counts touches for the nine fields included in this package.

```{r eval=FALSE}
library(phenoptr)
library(tidyverse)

# Find cell seg data files
base_path <- system.file("extdata", "samples", package = "phenoptrExamples")
files <- list_cell_seg_files(base_path)

# The phenotype pairs to locate. This will find CD8 cells touching
# tumor cells, and, separately, CD8 cells touching CD68 cells.
pairs <- list(c("CD8+", "CK+"),
             c("CD8+", "CD68+"))

# Colors for all the phenotypes mentioned in pairs
colors <- list(
  'CD8+' = 'yellow',
  'CK+' = 'cyan',
  'CD68+' = 'magenta'
)

# Set this true to write images in the same directory as the data files
write_images <- FALSE

# Count touching cells
touch_counts <- purrr::map_df(files, function(path) {
  count_touching_cells(path, pairs, colors, write_images=write_images)
})

glimpse(touch_counts)
```

```{r echo=FALSE}
# The previous is really slow, we fake it by saving the results...
touch_counts <- read_csv('touch_counts.csv')
glimpse(touch_counts)
```

----

## Aggregate per slide

The counts are easily aggregated over `Slide ID` using `dplyr::group_by` and 
`dplyr::summarize_at`:

```{r}
touch_counts %>% group_by(slide_id, phenotype1, phenotype2) %>% 
  summarize_at(vars(total1:touch_pairs), sum)
```

